---
output: 
  html_document: 
    keep_md: yes
    fig_caption:  yes
    toc: true
    toc_float: true
    
---




# Get ready for the workshop

This is an introduction to basic stats in R with a focus on regression. We will assume a familiarity with R to the extent of the participant having tried to do something with the programming environment in the past (e.g., a t-test), but no more than this. 

### R from zero warmup

If you have never used R in any way at a all, it may be helpful to take a quick look at this interactive tutorial for undergraduates at the University of Waterloo. Just complete the tutorial entirely online rather than downloading, starting from "R as a calculator".

[https://shiny.math.uwaterloo.ca/biology/QBshiny/IntroR/](https://shiny.math.uwaterloo.ca/biology/QBshiny/IntroR/)

Then, when you are ready, get some help installing R on your laptop. You will need R installed to participate in workshop. 


### Download needed files and packages

We will be using a flat .xlsx file of Hamilton Harbour data for our workshop exercise today. Our first job at the workshop will be to import that data. 

Please download the .xlxs to your personal computer prior to the workshop. 

In addition, we encourage you to download and install the packages we will be using for the workshop.

### Package to read .xlsx files

R is an open source platform, which means that anyone can contribute code for everyone to use. There are thousands of functions in R written by scientists, statisticians and mathematicians all over the world. Sometimes a group of functions related to a particular task are grouped together in what is called a “library” or "package". We are going to use one such library that has been created to import .xlsx files, "readxl". 

On your own computer, you will need download the package and install it. This is pretty easy. On the RStudio menu select *Tools*, and then *Install packages* from the drop down list. Enter *readxl* in the search window, and then select *Install*. You can also use the function install.packages. Just type *install.packages("readxl")*.


```{r, eval=FALSE}
install.packages("readxl")
```

The function **install.package()** downloads the indicated package from the internet and stores it. Eventually, in the console window you should see an indication the R has downloaded and installed the package. 

Once you've done this once, you don't need to do it ever again, unless you update the version of R you are using, or wish to use an updated version of the package.

In addition, the exercise on multivariate analysis will use the "vegan". Please install this package prior to the workshop as well.

These packages are now part of R’s set of functions, but since it they not part of the standard environment, you will have to mention to R when you wish to use functions from this collection. The **library()** function tells R you wish to use commands from a particular package on this occasion. 

Every time you restart R, or try to compile a .Rmd file, you will have to mention to R when you wish to use functions from this collection. The simplest method to do this is to always include the statement *library(readxl)* at the beginning of code that uses functions from this package.

Here R loads the package into active memory.

```{r, echo=TRUE}
library(readxl)
```

If this command works, you are ready to go for the first part of the workshop!

