---
output: 
  html_document: 
    keep_md: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic introduction to R

## Get ready for the workshop

This is an introduction to basic stats in R with a focus on regression. We will assume a familiarity with R to the extent of the participant having tried to do something with the programming environment in the past (e.g., a t-test), but no more than this. 

### R from zero warmup

If you have never used R in any way at a all, it may be helpful to take a quick look at this interactive tutorial for undergraduates at the University of Waterloo. Just complete the tutorial entirely online rather than downloading, starting from "R as a calculator".

[https://shiny.math.uwaterloo.ca/biology/QBshiny/IntroR/](https://shiny.math.uwaterloo.ca/biology/QBshiny/IntroR/)

Then, when you are ready, get some help installing R on your laptop. You will need R installed to participate in workshop. 


### Download needed files and packages

We will be using a flat .xlsx file of Hamilton Harbour data for our wrokshop exercise today. Our first job at the workshop will be to import that data. 

Please download the .xlxs to your personal computer prior to the workshop. 

In addition, we encourage you to download and install the packages we will be using.

### Package to read .xlsx files
We can import .csv files without any special tools (and we encourage you to consider this data format). So while base R includes the function **read.csv()** to read .csv files, to import proprietary formats like .xlsx we will need to use a _package_ to import the data. 

R is an open source platform, which means that anyone can contribute code for everyone to use. There are thousands of functions in R written by scientists, statisticians and mathematicians all over the world. Sometimes a group of functions related to a particular task are grouped together in what is called a “library” or "package". We are going to use one such library that has been created to import .xlsx files, "readxl". 

On your own computer, you will need download the package and install it. This is pretty easy. On the RStudio menu select 'Tools', and then 'Install packages' from the drop down list. Enter 'readxl' in the search window, and then select  'Install'. You can also use the function install.packages. Just type install.packages("readxl").


```{r, eval=FALSE}
install.packages("readxl")
```

The function install.package downloads the indicated package from the internet and stores it. Eventually, in the console window you should see an indication the R has downloaded and installed the package. 

Once you've done this once, you don't need to do it ever again, unless you update the version of R you are using, or wish to use an updated version of the package.

The package is now part of R’s set of functions, but since it is not a standard part of the program, you will have to mention to R when you wish to use functions from this collection. The **library()** function tells R you wish to use commands from that package on this occasion. 

Every time you restart R, or try to compile a .Rmd file, you will have to mention to R when you wish to use functions from this collection. The simplest method to do this is to always include the statement 'library(readxl)' at the beginning of code that uses functions from this package.

Here R loads the package into active memory.

```{r, echo=TRUE}
library(readxl)
```

If this command works, you are ready to go for the first part of the workshop!


## Workshop activities

### Data import & packages

Our first job today is to use a package to import the data we will be using throughout the workshop.

If you have sucessfully installed the library *readxl*, we can use it to read in the .xlsx file. The function we need is **read_excel()**. In general, I advise uses to examine the help functions in R every time they use a new function. If you type '?read_xlsx' in the console window, a help window will be called.

The help file indicates the inputs that the function takes. We see that we need the *path* to the .xlsx file, the *sheet* of that file we want to read. Note that if we do not specifiy which sheet, the function will default to reading the first sheet. Most functions in R will have default settings for the inputs so that you do not necessary have to set these. For example, to read in you Hamilton Harbour data, we are not going to using the *range* inpput that selects a cell range to read in. 

### Where’s my file??

There are at least three ways to find .xlsx file in R:

1. If you don’t know where you have saved a file (the *path*) you can combine the **file.choose()** function to get a drop down menu of your computers file system as read_excel(file.choose()). You can then use the menu to choose the appropriate file. </break> To import the data and save it to the variable name 'ham', enter the command 'ham = read_excel(file.choose())' in the Console window. This command opens a directory window and allows you to choose the file you wish to open.

2. Alternately, if you knew the location of your file, you could enter a command like ham = read_excel("/Users/Kim/Documents/hamiltonharbour.xlsx") to open the file immediately.

3. You could also simply change the directory that R is looking at. The command **getwd()** allows you to see which directory R is currently focusing on, and the command "setwd("C:/Users yourusername/workshop")" or similar should change the directory to the location you want to save to. You can check by typing getwd() again. Then you could simply enter "read_excel("ind dates flat all parameters Hamilton only FOR Kim.xlsx"), since R would already be focusing on the correct directory





```{r}
# xlsx files
ham <- read_excel("ind dates flat all parameters Hamilton only FOR Kim.xlsx")
```


### Did the import work??

Once you have read in the data, it is critically import to check that th import worked properly. Problems that can arise from incorrectly formatted files can then be resolved. Problems can include: numierc data being read in as character data, column names being read in as the the first row of data, missing values being coded as something other than NA.

Methods to examine the imported data: 

1. The easiest method is of course just to print the data. You can just type 'ham' in the Console window. However, most datasets will be too large for this approach to be useful (the top of the data will scroll right off the page). The **view()** function brings up a flat file view of the data which can be easier to use, but it is still pretty unwieldly for a dataset this size.

In that case, there are a number of commands to look at a portion of the dataset. 

2. The function **colnames(mydata)** will show you the all the column headings.  
3. The functions **head(mydata)** and **tail()** will show the first or last rows of the data.
4. The command **str(mydata)**. gives a whole bunch of information about your dataset including: the number of observations, the number of columns, the names of the different columns, the number of different string values in each text column (e.g., 7 levels for days of the week), the kind of values in the numeric columns (e.g., “int” for integer values). You should be pretty confident that your data was imported into R correctly after looking at all this.

```{r}
#View(ham)
colnames(ham)
tail(ham)
str(ham)
```

### Data in R

You’ll notice that the data are structured in columns. This is a *dataframe*, one of the most used data structures in R. Data structures are sets of variables organized in a particular way. In R there are 4 primary data structures we will use repeatedly.

*Vectors* are one-dimensional ordered sets composed of a single data type. Data types include integers, real numbers, and strings (character variables)

*Matrices* are two-dimensional ordered sets composed of a single data type, equivalent to the concept of matrix in linear algebra.

*Dataframes* are one to multi-dimensional sets with a row-column structure, and can be composed of different data types (although all data in a single column must be of the same type). In addition, each column in a data frame may be given a label or name to identify it. Data frames are equivalent to a flat file database, similar to spreadsheets (e.g., like a single excel spreadsheet).

*Lists* are compound objects of associated data. Like data frames, they need not contain only a single data type, but can include strings (character variables), numeric variables, and even such things as matrices and data frames. In contrast to data frames, list items do not have a row-column structure, and items need not be the same length; some can be a single value, and others a matrix. You can think of a list as a named box to put related objects into.

### Selecting portions of a dataframe
You can always grab just part of a dataframe. Dataframes are indexed by rows and columns. If you want the item from the 5th row and 2nd column type 'mydata[5,2]'. If you need just one column you can type either 'mydata[,2]', which grabs everything in column 2, or if you know the name of the column 'mydata\$Population'. To get rows 2 to 5 you can enter 'mydata[2:5,2]'. You can also combine this with the column name which may be easier to read, as 'mydata$Population[2:5]'.

Try it now: select just the first 15 rows of the dataframe you read into R.


```{r}
# by column
head(ham$Station_Acronym)

# by condition
ham$calanoid[ham$Station_Acronym=="HH6" & ham$year==2016]
ham$calanoid[ham$Station_Acronym=="HH6" & ham$year==2015]

```
### Data types in R

### Plotting data

```{r}
ham$area_group=as.factor(ham$area_group)
plot(calanoid~SamplingDate, data=ham, col=area_group)
legend("topleft", legend=levels(ham$area_group), 
       pch=1,col=c(1:length(levels(ham$area_group))))
```



### Summarizing data 

```{r}
table(ham$Station_Acronym)

knitr::kable(table(ham$Station_Acronym), col.names = c("Station Name", "No. of samples"))
```



### Plotting data

```{r}
samps=as.data.frame(table(ham[,c("year","area_group")]))
samps$year=as.numeric(as.character(samps$year))
plot(Freq~year,data=samps[samps$area_group=="deep",],
     xlim=c(min(samps$year), max(samps$year)),
     ylim=c(min(samps$Freq), max(samps$Freq)),
     col="black", type="b", main="Samples per year by area group")
lines(Freq~year,data=samps[samps$area_group=="NE",],
      col="red", type="b")
lines(Freq~year,data=samps[samps$area_group=="west",],
      col="blue", type="b")
lines(Freq~year,data=samps[samps$area_group=="wind",],
      col="green", type="b")

legend("topleft", legend=levels(ham$area_group), lty=1,
       pch=1,col=c("black", "red", "blue", "green"))
```

```{r}
# similar plot using ggplot
library(ggplot2)
ggplot(samps, aes(x = year, y = Freq, 
    group = area_group, colour = area_group)) + geom_line()

```


### Simple tests

